\Large Last Writer Slices and Communication Traps

### 2013-11-05 paper.tex
committing some changes to the debugging evaluation section, including two new figures.  also added some new changes to the introduction and the section on last writer slices.

We design efficient system support for collecting {\em last writer slices} for
shared-memory concurrent programs.  Last writer slices dynamically track memory
updates.  Last writer slices encode the provenance of values in memory, helping
with debugging.   We build on last writer slices, developing {\em communication
traps} (\ctraps).  \ctraps uses system support to efficiently monitor and
interpose on inter-thread memory dependences.  \ctraps provides an extensible
framework, exposing inter-thread dependences to {\em \ctraps applications}.
These can implement arbitrary dependence analyses.

We implement last writer slicing and \ctraps.  We show \ctraps is useful by
implementing two applications from prior work. We show that last writer slices
help with debugging using case studies of real-world bugs.  We evaluate our
designs on a set of server programs and standard benchmarks.  Our results show
\ctraps imposes overheads low enough for use in production systems (0--15\%)
running important services, like memcached, LevelDB, MySQL, and Apache.

@@

Debugging a concurrent program requires understanding how threads interact and
dependence information furnishes that understanding.  Dependence information
helps programmers with the difficult task of identifying control- and
data-flows that influenced a particular data value at a particular point.
Programmers debug programs more effectively when they can ask questions about
the provenance of values during a program's
execution~\cite{slicingsurvey,whylineicse}.  In concurrent programs, data
provenance information is particularly valuable because data may originate from
any of a potentially large number of threads, making reasoning complex.

@@

Our goal in this work is to collect, with deployable efficiency, useful
dependence information that captures data provenance for use in debugging and
serves as a foundation for more sophisticated program analyses.  We start from
two simple observations: (1) Reads are frequent so we must minimize tracking
overheads on reads, and (2) In concurrency analyses, {\em inter-thread}
dependences are most important.  Based on these observations, we focus first on
tracking write-after-write dependences (WAWs), which reflect data provenance.
Building on that, we describe support for tracking {\em communication
dependences}.  Communication occurs when a thread reads or overwrites a memory
location that was most recently written by a different thread; {\em i.e.},
inter-thread read-after-write dependences (RAWs) as well as (WAWs).  In this
work, we demonstrate that this form of light-weight dependence tracking is
useful and versatile, and has inherent overheads that are low enough for use in
deployed systems.  There are two parts to our work:

First, we describe {\em last writer slices}
(Section~\ref{sec:lastwriterslices}).  Last writer slices record the thread and
program point that last wrote each memory location.     Last writer slices are
useful because they expose the {\em provenance} of a memory location's contents
during debugging (Section~\ref{sec:debugging}).  When a failure occurs, last
writer slices are saved with a core dump.  Last writer slices are also
available during an execution in the debugger.  Note that while prior work
({\em e.g.}, ~\cite{recon,fasttrack,conmem,conseq,cci}) has relied on {\em ad
hoc} last-writer analyses, in this work, we isolate the essential last-writer
information in the form of last writer slices and provide an implementation
with overheads appropriate for production systems.  Tracking last writer slices
has low overhead because it imposes overhead on only write operations:  reads
are overhead-free.  In Section~\ref{sec:eval} we experimentally show that
collecting last writer slices indeed has overheads suitable for production
systems for a variety of applications.

@@

\paragraph{Non-overlapping Memory References} 
We assume each written memory location corresponds to a non-overlapping memory
region because our \lwt design is {\em granularity oblivious}.  We discuss this
issue only informally due to space constraints.  Each \lwt entry corresponds to
a memory location referenced by a program instruction.  Two instructions might
reference different locations that actually overlap because the referenced
locations have different granularities.  When that happens, both accesses
update different \lwt entries, because they reference different locations.
However, the overlapping region then has two different, inconsistent entries
reflecting its last {\em two} updates, rather than just its last write.

For example, one instruction might write a word and another might later write
the second byte of that word.  The word access updates the word's \lwt entry.
The byte access updates a second \lwt entry for the byte address.  However, the
byte is part of the previously written word, meaning the \lwt entry for the
word no longer reflects reality -- the word was last written by the instruction
and thread that wrote the byte, not the one that wrote the word.  This
situation leads to incompleteness because the entry for the word should be
updated but it is not.

Assuming non-overlapping regions ensures \lwt are correct.  Another
approach would be to assume a fixed \lwt granularity.  We chose our solution
for two reasons.  First, fixing the \lwt at a large granularity ({\em i.e.},
cache-line) precludes sound slicing for fine-grained accesses.  Second, fixing
the \lwt at a small granularity ({\em i.e.}, byte) requires multiple \lwt
updates for larger granularity accesses, increasing overheads.

@@

\paragraph{Case Study: HTTrack}
The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
and the text in Section~\ref{sec:httdebug} describes the benefit provided by
last writer slices in debugging an ordering violation in HTTrack-3.43.9.

\paragraph{Case Study: PBZip2}
We conducted a case study using a buggy version of PBZip2-0.9.1.    The bug is
a use-after-delete bug that leads to an assertion failure.  A worker thread
tries to acquire a mutex after it has been deleted by the main thread during
shutdown, causing the assertion to fail.

@@

\paragraph{Case Study: MySQL}
Last writer slices were also useful for debugging a logging system failure in
MySQL-4.0.12.  The bug is an atomicity violation involving two threads.  One
thread rotates the database's log file.  Log rotation involves closing the log
file, marking the log closed, opening a new log file, and marking the log open.
Log rotation should be atomic, but lack of synchronization allows another
thread to run queries during rotation.  The query code checks if the log is
marked closed; if it is not closed, the queries are logged, but if it is
closed, the queries are not logged and the program simply continues.  We added
an assertion to the non-logging case because this behavior is a security
vulnerability.  The bug report describes the bug as ``critical'', so it is
reasonable to assume a programmer that expected all queries to be logged would
write such an assertion after the bug was reported.

We ran the program and concurrently issued a log rotation command, and an
insert query.  The bug manifested ({\em i.e.}, the queries went unlogged) and
our assertion failed, dumping core.  Examining the log's last writer slice in
GDB allowed examining the last thread to mark the log closed.  We discerned it
was the log rotation thread that last marked the log closed during log
rotation.  Knowing how the log ended up marked closed enables reasoning about
what code to change ({\em e.g.}, added synchronization) to prevent the
situation in the future.

While a developer is likely to understand the symptom of the bug (queries go
unlogged) without the information in the \lwt, understanding {\em why} it is marked closed would be
difficult.  This example illustrates how the \lwt fills in the crucial inter-thread
dependence that helps understand why the program failed.

\subsection{Setup and Benchmarks}

@@

\subsection{Performance}
\label{sec:eval:perf}

>>>>

We design efficient system support for collecting {\em last writer slices} for
shared-memory concurrent programs.  Last writer slices dynamically track memory
updates.  Last writer slices reflect the provenance of values in memory,
information that is useful during debugging.   Using last writer slices,
we build {\em communication traps} (\ctraps).  \ctraps is a mechanism that
efficiently monitors and can interpose on memory operations that share data
between threads.  \ctraps is an extensible framework, exposing
inter-thread dependences to {\em \ctraps applications} that can implement
arbitrary dependence analyses.

We implement last writer slicing and \ctraps.  We show that last writer slices
provide information that is useful for debugging several real-world bugs.  We
show \ctraps is useful by implementing two concurrency analyses from prior
work.  We evaluate our designs on a set of server programs and standard
benchmarks.  Our results show that collecting last writer slices has low run
time overheads (0--15\%) for many important applications, like memcached,
LevelDB, MySQL, and Apache.  We show that \ctraps imposes low run time
overheads by default and that the overheads scale with the complexity of the
analysis being implemented.

@@

Debugging a concurrent program requires understanding how threads interact and
dependence information furnishes that understanding.  Dependence information
helps programmers with the difficult task of reasoning about the {\em
provenance} of a memory location's value -- control- and data-flows that
influenced that value.  Programmers debug programs more effectively when they
have access to a value's provenance at points during a program's
execution~\cite{tipslicingsurvey,whylineicse}.  In concurrent programs, data
provenance information is particularly valuable because data may originate from
any of a potentially large number of threads, making reasoning without explicit
provenance information complex and difficult.

@@

Our goal in this work is to collect, with deployable efficiency, useful
dependence information that captures data provenance for use in debugging and
serves as a foundation for more sophisticated program analyses.  We start from
two simple observations: (1) Reads are frequent so we must minimize tracking
overheads on reads, and (2) In concurrency analyses, {\em inter-thread}
dependences are most important.  Based on these observations, we focus first on
tracking write-after-write dependences (WAWs), which provide a useful form of
data provenance information.  Building on that, we describe support for
tracking {\em communication dependences}.  Communication occurs when a thread
reads or overwrites a memory location that was most recently written by a
different thread; {\em i.e.}, inter-thread read-after-write dependences (RAWs)
as well as WAWs.  In this work, we demonstrate that this form of light-weight
dependence tracking is useful and versatile, and has inherent overheads that
are low enough for use in deployed systems.  There are two parts to our work:

First, we describe {\em last writer slices}
(Section~\ref{sec:lastwriterslices}).  Last writer slices record the thread and
program point that last wrote each memory location.     Last writer slices are
useful because they expose the provenance of a memory location's contents
during debugging (Section~\ref{sec:debugging}).  When a failure occurs, last
writer slices are saved with a core dump.  Last writer slices are also
available during an execution in the debugger ({\em e.g.}, at breakpoints).
Note that while prior work ({\em e.g.},
~\cite{recon,fasttrack,conmem,conseq,cci}) has relied on {\em ad hoc}
last-writer analyses, in this work, we isolate the essential last-writer
information in the form of last writer slices and provide an implementation
with overheads appropriate for production systems.  Tracking last writer slices
has low overhead because it imposes overhead on only write operations:  reads
are overhead-free.  In Section~\ref{sec:eval} we experimentally show that
collecting last writer slices indeed has overheads suitable for production
systems for a variety of applications.

@@

\paragraph{Non-overlapping Memory References} 
We assume each written memory location corresponds to a non-overlapping memory
region because our \lwt design is {\em granularity oblivious}.  We discuss this
issue informally due to space constraints.  Each \lwt entry corresponds to a
memory location referenced by a program instruction.  Two instructions might
reference different locations that overlap because the referenced locations
have different granularities ({\em e.g.}, accessing a byte that is part of a
previously accessed word).  When that happens, both accesses update different
\lwt entries, because they reference different locations.  However, the
overlapping region then has two different entries reflecting the last {\em two}
updates to its parts rather than just its last write.

The \lwt remains correct, assuming the program only accesses non-overlapping
regions.  In programs that access overlapping regions, there is always some
entry in the \lwt that correctly reflects each memory location's last writer.
However, the entry for a coarse-grained region may be obsolete because of an
access to an overlapping sub-region.

One approach to mitigating this issue would be to allocate \lwt entries for
regions of fixed granularity.  We chose not to do so for two reasons.  First,
associating \lwt entries with regions of  large granularity only ({\em i.e.},
cache-line) precludes sound analysis for fine-grained accesses.  Second,
associating \lwt entries with regions of small granularity ({\em i.e.}, byte)
requires multiple \lwt updates for larger granularity accesses, increasing
their overheads proportionately to the granularity of the access.

@@

\paragraph{HTTrack}
The running example in Figures~\ref{fig:coreDumpFail} and ~\ref{fig:httlws},
and the text in Section~\ref{sec:httdebug} describes the benefit provided by
last writer slices in debugging an ordering violation in HTTrack-3.43.9.

\paragraph{PBZip2}
We conducted a case study using a buggy version of PBZip2-0.9.1.    The bug is
a use-after-delete bug that leads to an assertion failure.  A worker thread
tries to acquire a mutex after it has been deleted by the main thread during
shutdown, causing the assertion to fail.

@@

\paragraph{MySQL}

We used last writer slices to debug an atomicity violation in MySQL-4.0.12.
Figure~\ref{fig:mysqllws} shows involved code.  Thread 1 rotates the database's
log file, marking it closed, opening a new log file, then marking it open
again.  Log rotation should be atomic, but the lack of synchronization allows
Thread 2 to run its insert during log rotation.  The insert code checks if the
log is marked closed. If it is insert is not logged, manifesting the failure.
We added an {\tt abort} statement to the non-logging case because this behavior
is a security vulnerability that the bug report describes as ``critical''.
Adding this {\tt abort} is reasonable, as the bug report describes the
condition in the {\tt else} block as being an error.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/MySQLDebug.pdf}
\caption{\label{fig:mysqllws}{\bf Using last writer slices to debug an
atomicity violation in MySQL 4.0.12.} The last writer slice reveals the code that closes the log before a database insertion that goes unlogged, which is a security problem.}
\end{figure}

We triggered the bug by running an insert query during a log rotation.  The bug
manifested, stopping execution at point {\bf C}.  As described in the bug
report, the query went unlogged and examining the core dump revealed the log
was marked closed at point {\bf B}.  However, the core dump did not reveal {\em
why} the log was marked closed there.  Looking at the last writer slice for the
{\tt log\_type} variable shows that it was written last Thread 1 at {\bf A}.
Seeing here that {\tt log\_type} is accessed by multiple threads
suggests adding synchronization to prevent this atomicity violation.

\paragraph{JDK-StringBuffer}

Last writer slices help elucidate a subtle atomicity violation in the
StringBuffer library in version 1.4 of the Java JDK.  We use the C++ port of
the buggy JDK and a driver program code developed in prior concurrency debugging
work~\cite{concurrencybugs}.

Figure~\ref{fig:jdklws} shows how the bug can manifest as a failure. The string
object pointed to by {\tt sb} in Thread 1's code contains a string buffer and a
variable called {\tt count} that represents its length.  The {\tt append()}
code should atomically read {\tt sb->count} and update the string's buffer.
The failure occurs when the {\tt append()} code in Thread 1 is interleaved with
the {\tt erase()} code in Thread 2 as indicated in the figure.  The {\tt
erase()} code updates the buffer's length, making the version Thread 1 read
stale.  Thread 1 uses its stale version in the function called at point {\bf
B}, causing a crash.

\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{figs/JDKStringBufferDebug.pdf}
\caption{\label{fig:jdklws}{\bf Using last writer slices to debug an
ordering violation in the JDK 1.4 StringBuffer library.} The last writer slice reveals the interaction between the erase function and the append function that leads to a failure.}
\end{figure}

When the atomicity violation manifests, the appending thread is updating the
string's underlying buffer.  The erasing thread, however, may have completed
its erase operation.  The execution context of the erasing thread at the point
of the failure provides no insight as to why the program failed.  Examining the memory image, the programmer can only conclude that the length and buffer are 
inconsistent, but not why they are inconsistent.

This situation illustrates the value of the provenenace information furnished
by last writer slices.  At the point of the failure the last writer slice for
the length variable points to a decrement of that variable in the erase
function.  The last writer slice illustrates the need for synchronization
between the append function and the erase function.  In the absence of
provenance information, the programmer would not be aware of that relationship.

\subsection{Performance Evaluation: Setup and Benchmarks}

@@

\subsection{Performance Evaluation}
\label{sec:eval:perf}
